---
title: "Data Science Applied to Ag - Final Project - ML"
format:
  html:
    embed-resources: true
    toc: true
    theme: cerulean
author: Md Shakir Moazzem, Umar Munir
---

# Introduction  

This script contains the data wrangling steps for the final project.  

<<<<<<< HEAD
# Setup  

##Loading packages  

```{r Setup, message=F}

# Installing packages
install.packages("readxl")
install.packages("janitor")
install.packages("dplyr")
install.packages("tidyr")
install.packages("readr")
install.packages("lubridata")



# Loading packages (every time you open R)
library(readxl) # to read excel files
library(janitor) # to clean data; helps fix and standardize the column names
library(dplyr) # wrangling
library(tidyr) # wrangling
library(readr) # to export csv
library(lubridate)
library(stringr)

```


## Reading data  

The following code chunk is to read the csv files for the 3 training data sets

```{r training data import, message=F, warning=F}

#reading the csv files for the 3 training data sets 

trait <- read_csv("../data/training/training_trait.csv") 
meta  <- read_csv("../data/training/training_meta.csv")
soil  <- read_csv("../data/training/training_soil.csv")

# Checking first rows
trait
meta
soil

# Viewing the data frame in a separate tab
View(trait)
View(meta)
View(soil)

```

The following code chunk is to read the csv files for the 3 test data sets

```{r test data import, message=F, warning=F}

#reading the csv files for the 3 training data sets 

submission_test  <- read_csv("../data/testing/testing_submission.csv")
meta_test <- read_csv("../data/testing/testing_meta.csv") 
soil_test  <- read_csv("../data/testing/testing_soil.csv")

# Checking first rows
submission_test
meta_test #contains meta data for the year 2024 only
soil_test #contains soil data for the year 2024 only

# Viewing the data frame in a separate tab
View(submission_test)
View(meta_test)
View(soil_test)


```

# EDA 

The following code chunk is to conduct EDA (Exploratory Data Analysis) for the training data sets

```{r EDA, message=F, warning=F}

#EDA for "training_trait.csv"
summary(trait)

#EDA for "training_meta.csv"
summary(meta)

#EDA for "training_soil.csv"
summary(soil)

```

## Data wrangling 

The following code chunk is to conduct data wrangling on the the training data sets

```{r data wrangling, message=F, warning=F}

trait_w <- trait %>%
  #standardizing the column names
  clean_names() %>%
  #adjusting yield to a 15.5% moisture basis; Conversion formula: yield_15.5 = (yield * (1 - M/100)) / (1 - 0.155)
  mutate(yield_15.5 = (yield_mg_ha * (1 - grain_moisture / 100)) / (1 - 0.155)) %>%
  #Feature engineering corn growth stages based on each growth‐state window using planting and harvest date  
  mutate(
    planting_date = as.Date(date_planted, "%m/%j/%y"),
    harvest_date = as.Date(date_harvested, "%m/%j/%y"),
    plant_doy = yday(planting_date),
    harvest_doy = yday(harvest_date),
    season_len = harvest_doy - plant_doy
    )
  
summary(trait_w)

View(trait_w)


```


```{r}

soil_w <- soil %>%
  separate(
    col   = site,
    into  = c("site_left", "site_right"),
    #splitting based on the last underscore (i.e. second underscore where present)
    sep   = "_(?=[^_]*$)",
    remove = TRUE,
    extra  = "merge",
    fill   = "right"
  ) %>%
  select(-site_right) %>%
  rename(site = site_left)

# View the result
View(soil_w)
=======
#Code Setup

```{r}
library(tidyverse) # data wranggling and plotting
library(car) # Anova functions
library(emmeans) # model residuals extraction
library(multcomp) # pairwise comparison letter display
library(broom)
library(readxl)
library(dplyr)
```

#Data load

```{r, meta}
cvt_meta <- read.csv("../data/training/training_meta.csv") 

view(cvt_meta)

```

```{r, soil}
cvt_soil <- read.csv("../data/training/training_soil.csv") 

nrow(cvt_soil)
```

```{r, trait}
cvt_trait <- read.csv("../data/training/training_trait.csv") 

nrow(cvt_trait)
```


#Data exploration

```{r, meta}
summary(cvt_meta)
```


```{r, soil}
summary(cvt_soil)
#This one is missing year 2014 data for the soil while the other two have it.
```


```{r, trait}
summary(cvt_trait)
```

#Cvt_Soil_wrangling

```{r}
cvt_soil_cmpt <- cvt_soil %>% 
  bind_rows(data.frame(year = 2014, value1 = NA )) %>% 
  arrange(year)

cvt_soil_cmpt
```


#Combine trait and soil data

```{r}

#All the useful variables are in these two datasets. I believe the meta data is only helpful if we want to pull more open data variable therefore I am going ahead and combining these two. Then I will convert the moisture percentage to 15.5.

cvt_comb <- left_join(cvt_trait, cvt_soil, by = "site") 

cvt_comb

view(cvt_comb)
```


#Conversion of yield to 15.5% moisture rate

```{r}
>>>>>>> 836dfb6f11c8fc4c40941b682918ce1763c0909d

```















<<<<<<< HEAD










































=======
>>>>>>> 836dfb6f11c8fc4c40941b682918ce1763c0909d
